Rethinking Felony Disenfranchisement: The Case of New York City and Miami

### 2019-05-14 markdown/kmorris_thesis.Rmd
start coding zipcodes keep writing

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the *state* level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, less work has been done exploring the types of neighborhoods in which these effects are concentrated, and the impact these laws might have at the local level. @Burch2013 is an exception to this. In this paper, Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block-group level turnout and involvement with the criminal justice system. She determines that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods." This paper seeks to expand on her work by replicating her findings in New York City and by introducing a new way of

@@

##Identification of Lost Voters {-}
Because I am interested in the spatial effects of felony disenfranchisement at the local level, I explore the impact of disenfranchisement on turnout in the most recent mayoral election, which occured on November 7^th^, 2017. I begin by indentifying "lost voters," defined as individuals who were ineligible to cast a ballot in the 2017 (either because of incarceration of parole status) and had cast a ballot between 2007 and 2016. Lost voters are of particular interest because of their demonstrated propensity to vote. I assume that disenfranchisement is likely to have a bigger impact where it directly impacts residents with a relatively recent history of casting a ballot.

These individuals are identified by matching Department of Corrections and Community Supervisions data to the voter file using the methodology developed by Burch [-@Burch2010; -@Burch2011]. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I report estimates of the potential impact of these false positives in the Appendix.] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison (and city council districts). There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

```{r citywide-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:citymap}Lost Voters on Election Day, 2017", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./output/city_map.RDS")
print(j)
```
The spatial concentration of lost

##Geocoding {-}

#Maps {-}

As we can see in Figure \ref{fig:citymap}, all these other observations

#Match Output {-}
```{r block-table-chunk, echo = F}
t <- readRDS("./temp/match_table_block_group.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:blockgroup-table) Results of Block Group-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```

```{r tract-table-chunk, echo = F}
t <- readRDS("./temp/match_table_tract.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:tract-table) Results of Tract-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```
At both the tract and block group level, matching results in an untreated group of neighborhoods that looks substantially like our treatment group. As discussed above, the types of neighborhoods targeted by felony disenfranchisement look far different than the average New York City neighborhood. They have far lower incomes and are much less White; they also have lower median ages and more residents are registered as Democrats.

Reg table
```{r tract-reg-chunk, echo = F, results = "asis", message = F}
tract_reg <- readRDS("./temp/match_reg_tract.rds")
tract_reg_black <- readRDS("./temp/match_reg_tract_nhb.rds")
bg_reg <- readRDS("./temp/match_reg_block_group.rds")
bg_reg_black <- readRDS("./temp/match_reg_block_group_nhb.rds")

stargazer(bg_reg, bg_reg_black, tract_reg, tract_reg_black,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Block Group Level", "Tract Level", "Tract Level"),
          header = F,
          type = "latex",
          covariate.labels = c("D(Neighborhood Lost a Voter)", "Share Non-Hispanic Black", "D(Lost Voter) * Share Black", "Constant"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:tract-reg) Matching Regression",
          table.placement = "H",
          style = "ajps",
          omit.stat = c("f", "ser"))
```
Although the removal of a voter is not generally associated with decreased turnout, we *do* see an impact on turnout in the block groups and census tracts

>>>>

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the *state* level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, less work has been done exploring the types of neighborhoods in which these effects are concentrated, and the impact these laws might have at the local level. @Burch2013 is an exception to this. In this paper, Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block-group level turnout and involvement with the criminal justice system. She determines that "at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods." This paper seeks to expand on her work by replicating her findings in New York City and by introducing a new way of measuring the impact.

@@

##Identification of Lost Voters {-}
In this analysis, I offer a slightly different defintion of "lost voter" than much of the literature. Many recent papers have attempted to identify relationships between the number of *potentially* lost voters and turnout. Here, I explore whether the disenfranchisement of voters with a history of participating in elections is related to neighborhood turnout. In this analysis, "lost voters" are individuals ineligible to cast a ballot on a given election day who have cast a ballot in the previous ten years. To study the effect of felony disenfranchisement on voting at the local level (and its potential implications for the distribution of political power at the local level), I use turnout in the most recent election for city-wide office in New York - the mayoral election which took place on November 7^th^, 2017. Lost voters, therefore, are all individuals who were incarcerated or on parole on November 7^th^, 2017, and had cast a ballot between 2007 and 2016.

These individuals are identified by matching Department of Corrections and Community Supervisions data to the voter file using the methodology developed by Burch [-@Burch2010; -@Burch2011]. I use first, middle, and last names, and dates of birth, to join these two datasets.^[Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I report estimates of the potential impact of these false positives in the Appendix.] Figure \ref{fig:citymap} shows where these lost voters lived before going to prison, with city council districts also included. There were `r comma(readRDS("./temp/lost_count_nyc.rds"))` such lost voters within New York City as of the 2017 general election, and `r comma(nrow(readRDS("./temp/ids_of_lost_voters.rds")))` statewide.

```{r citywide-map, echo = FALSE, message = F, warning = F, fig.width = 5, fig.height = 5, fig.cap = "\\label{fig:citymap}Lost Voters on Election Day, 2017", out.extra = '', fig.pos = "H", fig.align = "center"}
j <- readRDS("./output/city_map.RDS")
print(j)
```
The spatial concentration of lost is readily apparent. In some communities, such as Greenwich Village and Brooklyn Heights, hardly any voters were disqualified from participating in the 2017 elections. In other communities, such as Harlem and Central Brooklyn, large numbers of individuals were not allowed to cast a ballot for mayor.

## Testing for Neighborhood Turnout Effects {-}

To test whether neighborhoods with lost voters had systematically lower turnout rates than similar neighborhoods without such disqualifications, I begin by using a matching model. Neighborhoods (defined as census tracts and block groups) are considered treated if they were home to *at least one* lost voter in the 2017 election; they are untreated if no residents were disqualified from the election because of a felony conviction. I use a genetic match to match treated to untreated census tracts and block groups. I match on a series of demographic and political indicators: racial characteristics, median income, education, age, population, share noncitizen,^[The Census Bureau does not make noncitizen estimates available at the block group level. As such, block groups are assigned their census tract's share noncitizen for matching purposes.] and partisan composition. Each treated census tract is matched to 10 untreated census tract, while each treated block group is treated to 30 untreated block groups. Matching is done with replacement. Tables \ref{tab:blockgroup-table} and \ref{tab:tract-table} present the results of these matches.

### Match Output {-}
```{r block-table-chunk, echo = F}
t <- readRDS("./temp/match_table_block_group.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:blockgroup-table) Results of Block Group-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```

```{r tract-table-chunk, echo = F}
t <- readRDS("./temp/match_table_tract.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:tract-table) Results of Tract-Level Matching", linesep = "") %>% 
  add_header_above(c(" " = 1, "Means: Unmatched Data" = 2, "Means: Matched Data" = 2, "Percent Improvement" = 4), align = "c") %>% 
  kable_styling(latex_options = c("scale_down", "HOLD_position"))
j
```
At both the tract and block group level, matching results in an untreated group of neighborhoods that looks substantially like our treatment group. These tables also demonstrate the striking extent to which neighborhoods with lost voters differ from the average neighborhood in New York City. Neighborhoods with lost voters are substantially less white, have far lower median incomes, and a larger share of voters are registered as Democrats.

After matching neighborhoods, I use a simple regression to test whether turnout in the 2017 mayoral election was different in areas with lost voters. Census tract and block group level turnout rates are calculated using the geocoded voter file. Each voter's record indicates whether the voter participated in the 2017 general election,^[The New York State voter file is not without errors. For instance, the voter file erroneously does not record the author as participating in the 2017 mayoral election. However, insofar as these errors are not geographically correlated, there is little reason to believe they impact the results of this analysis. Unfortunately, testing the distribution of these errors is not possible.] which are then aggregated to estimate the number of ballots cast in each neighborhood. The number of ballots cast is divided by estimated voting age population in the neighborhood as reported by the Census Bureau.^[Voting age population is of course only an approximation of the number of eligible voters; notably, its use will underestimate turnout in neighborhoods with large noncitizen populations. However, treatment and matched control neighborhoods have comparable rates of noncitizenship, lessening any potential bias.]

Much of the literature has discussed whether felony disenfranchisement is particularly demobilizing for eligible Black voters. I therefore include models which explore any potential difference in treatment effect in neighborhoods where a higher share of the population is Black.

Table \ref{tab:match-reg} presents the results of these regression models.
```{r tract-reg-chunk, echo = F, results = "asis", message = F}
tract_reg <- readRDS("./temp/match_reg_tract.rds")
tract_reg_black <- readRDS("./temp/match_reg_tract_nhb.rds")
bg_reg <- readRDS("./temp/match_reg_block_group.rds")
bg_reg_black <- readRDS("./temp/match_reg_block_group_nhb.rds")

stargazer(bg_reg, bg_reg_black, tract_reg, tract_reg_black,
          column.separate = c(2, 2),
          column.labels = c("Block Group Level", "Tract Level"),
          header = F,
          type = "latex",
          covariate.labels = c("D(Neighborhood Lost a Voter)", "Share Non-Hispanic Black", "D(Lost Voter) * Share Black", "Constant"),
          dep.var.labels = c("Turnout Rate"),
          title = "(\\#tab:match-reg) Matching Regression",
          table.placement = "H",
          style = "ajps",
          omit.stat = c("f", "ser"))
```
Although the removal of a voter is not generally associated with decreased turnout, we *do* see an impact on turnout in the block groups and census tracts
